---
title: "Lichess Game Analysis"
author: "Steven Sparber"
date: '2023-01-06'
output: html_document
---

## Reading in Data

```{r}
library(dplyr)
library(stringr)

games <- read.table("Chess Games (6.1.21 - 11.30.22).pgn", sep="\n")

str(games)
```

## Data Cleaning

#### First steps in cleaning and creating data frame

```{r}

# Removal of brackets

games$V1 = gsub("\\[|\\]", "", games$V1)

# Removal of rows with unnecessary information

games = games %>%
  filter(!grepl('https|UTC|ECO|Title', V1))

# Splitting the one long column into 14 separate ones based on the row number where information repeats

games = as.data.frame(split(games, 1:14))

colnames(games) = c("Game_Type", "Date", "White", "Black", "Result", "White_Rating", "Black_Rating", "White_Rating_Diff", "Black_Rating_Diff", "Variant", "Time_Control", "Opening", "Termination", "Moves")

# Creating a backwards ID column since the games are in reverse chronological order:

games$ID = nrow(games):1

```

#### Skimming the data frame

```{r}

library(skimr)

skim(games)

```

#### Cleaning the Game_Type column

```{r}
## Seeing all values in Game_Type column

table(games$Game_Type)

## Removing unnecessary words from Game_Type column and converting to a factor variable

games$Game_Type = gsub("Event | game", "", games$Game_Type)
games$Game_Type = as.factor(games$Game_Type)

## Creating binary indicator columns for game type

games$Hourly_Blitz_Arena = as.numeric(games$Game_Type == "Hourly Blitz Arena")
games$Hourly_SuperBlitz_Arena = as.numeric(games$Game_Type == "Hourly SuperBlitz Arena")
games$Rated_Blitz = as.numeric(games$Game_Type == "Rated Blitz")
games$Rated_Bullet = as.numeric(games$Game_Type == "Rated Bullet")
games$Rated_Rapid = as.numeric(games$Game_Type == "Rated Rapid")

## Viewing cleaned columns

games %>%
  select(Game_Type, Hourly_Blitz_Arena, Hourly_SuperBlitz_Arena, Rated_Blitz, Rated_Bullet, Rated_Bullet, Rated_Rapid)
```

#### Cleaning Date column

```{r}
## Removing unnecessary words from Date column and converting to a date variable

games$Date = gsub("Date ", "", games$Date)
games$Date = as.Date(str_replace_all(games$Date, '\\.', '/'))

## Creating "Weekday" column from "Date" column

games$Day = factor(weekdays(games$Date), ordered = TRUE, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))

## Creating Month & Year column from "Date" column

library(zoo)
games$Month_Year = as.yearmon(games$Date, "%Y-%m") 

## Creating binary columns for weekdays

games$Monday = (as.numeric(games$Day == 'Monday'))
games$Tuesday = (as.numeric(games$Day == 'Tuesday'))
games$Wednesday = (as.numeric(games$Day == 'Wednesday'))
games$Thursday = (as.numeric(games$Day == 'Thursday'))
games$Friday = (as.numeric(games$Day == 'Friday'))
games$Saturday = (as.numeric(games$Day == 'Saturday'))
games$Sunday = (as.numeric(games$Day == 'Sunday'))

## Viewing cleaned columns

games %>%
  select(Date, Month_Year, Day, Monday, Tuesday, Wednesday, Thursday, Friday, Saturday, Sunday)
```

#### Cleaning for columns: White, Black, Result, White_Rating, Black_Rating, White_Rating_Diff, Black_Rating_Diff

```{r}
## Removing unnecessary words from "White" and "Black" column

games$White = gsub("White ", "", games$White)
games$Black = gsub("Black ", "", games$Black)

## Removing unnecessary words from Result column

games$Result = gsub("Result ", "", games$Result)

## Removing unnecessary words from White_Rating and Black_Rating columns, then converting to a numeric variable

games$White_Rating = as.numeric(gsub("WhiteElo ", "", games$White_Rating))
games$Black_Rating = as.numeric(gsub("BlackElo ", "", games$Black_Rating))

## Removing unnecessary words from White_Rating_Diff and Black_Rating_Diff column, then converting to a numeric variable

games$White_Rating_Diff = as.numeric(gsub("WhiteRatingDiff ", "", games$White_Rating_Diff))
games$Black_Rating_Diff = as.numeric(gsub("BlackRatingDiff ", "", games$Black_Rating_Diff))

```

#### Removing Variant column

```{r}
table(games$Variant) #Only one, so we can remove this column

games = games %>%
  select(-Variant)
```

#### Cleaning Time_Control column

```{r}
## Inspecting Time_Control column and creating transformations

table(games$Time_Control) #Looks like this is in seconds. We'll also need to create an "increment" column. Also should remove, "TimeControl"

# Removing unnecessary words from Time_Control column

games$Time_Control = gsub("TimeControl ", "", games$Time_Control)

# Splitting Time_Control column into minutes and increment

library(tidyr)

games = separate(data = games, col = Time_Control, into = c("Time_Control_Minutes", "Time_Control_Increment"), sep = "\\+", remove = FALSE)

# Converting original Time_Control column into a factor variable and new Time_Control columns to numeric

games$Time_Control = as.factor(games$Time_Control)
games$Time_Control_Minutes = as.numeric(games$Time_Control_Minutes)
games$Time_Control_Increment = as.numeric(games$Time_Control_Increment)
```

#### Cleaning Opening column

```{r}
## Removing unnecessary words from Opening column and creating Opening_Type and Opening_Variation columns

games$Opening = gsub("Opening ", "", games$Opening)

games = games %>%
  separate(col = Opening, into = c("Opening_Type", "Opening_Variation"), sep = ": ", remove = FALSE)

## Converting all opening columns into factors

games$Opening = as.factor(games$Opening)
games$Opening_Type = as.factor(games$Opening_Type)
games$Opening_Variation = as.factor(games$Opening_Variation)

## Creating e4 and d4 opening columns

games$e4 = as.numeric(str_detect(games$Moves, '1. e4'))
games$d4 = as.numeric(str_detect(games$Moves, '1. d4'))

## Viewing new and cleaned columns

games %>%
  select(Opening, Opening_Type, Opening_Variation)
```

#### Cleaning Termination column

```{r}
## Removing unnecessary words from Termination column and converting to factor, note we keep all games

games$Termination = as.factor(gsub("Termination ", "", games$Termination))
```

#### Cleaning and creation of columns to show results

```{r}
## Creating columns to show whether I played as black or white

games = games %>%
  mutate(Played_White = if_else(White == 'username', 1, 0),
         Played_Black = if_else(Black == 'username', 1, 0))

## Separating white and black win columns into binary variables

games = separate(data = games, col = Result, into = c("White_Result", "Black_Result"), sep = "\\-")

# Assigning a value of 0.5 to draws and making it a numeric value

games$White_Result = as.numeric(gsub('1/2', '0.5', games$White_Result))
games$Black_Result = as.numeric(gsub('1/2', '0.5', games$Black_Result))

# Creating columns to show if I won or loss, overall and as black or white

games = games %>%
  mutate(I_Win_As_White = if_else(Played_White + White_Result == 2, 1, 0),
         I_Win_As_Black = if_else(Played_Black + Black_Result == 2, 1, 0),
         I_Lose_As_White = if_else(I_Win_As_White == 0, 1, 0),
         I_Lose_As_Black = if_else(I_Win_As_Black == 0, 1, 0),
         I_Draw_As_White = if_else(Played_White + White_Result == 1.5, 1, 0),
         I_Draw_As_Black = if_else(Played_Black + Black_Result == 1.5, 1, 0),
         My_Result = if_else(I_Win_As_White + I_Win_As_Black == 1, 1,
                             if_else(I_Draw_As_White + I_Draw_As_Black == 1 , 0.5, 0)),
         My_Result_Cat = as.factor(if_else(My_Result == 1, "Win",
                                           if_else(My_Result == 0.5, "Draw", "Loss"))),
         I_Win = if_else(My_Result == 1, 1, 0),
         I_Lose = if_else(My_Result == 0, 1, 0),
         I_Draw = if_else(My_Result == 0.5, 1, 0))

## Creating extra column to show if I was black or white

games = games %>%
  mutate(White_or_Black = as.factor(if_else(White == "username", "White", "Black")))

## Viewing new and cleaned columns

games %>%
  select(Played_White, Played_Black,
         White_Result, Black_Result,
         My_Result, My_Result_Cat,
         I_Win, I_Lose, I_Draw)

```

#### Creating columns to show total number of moves of major and minor pieces as well as king

```{r}

## Removing result from Moves column

games$Moves = str_remove_all(games$Moves, ' 1-0')
games$Moves = str_remove_all(games$Moves, ' 0-1')
games$Moves = str_remove_all(games$Moves, ' 1\\/2-1\\/2')

## Creating column for number of moves. We do this by counting and dividing spaces using the stringr library and str_count function.

games = games %>%
  mutate(Num_Moves = as.numeric(ceiling((str_count(Moves, " "))/3)))

## Creating columns for major and minor piece moves and King

games = games %>%
  mutate(White_Bishop_Moves = str_count(Moves, '\\. B'),
         White_Knight_Moves = str_count(Moves, '\\. N'),
         White_Rook_Moves = str_count(Moves, '\\. R'),
         White_King_Moves = str_count(Moves, '\\. K'),
         White_Queen_Moves = str_count(Moves, '\\. Q'),
         Bishop_Moves = str_count(Moves,' B'),
         Knight_Moves = str_count(Moves, ' N'),
         Rook_Moves = str_count(Moves, ' R'),
         King_Moves = str_count(Moves, ' K'),
         Queen_Moves = str_count(Moves, ' Q'),
         Black_Bishop_Moves = Bishop_Moves - White_Bishop_Moves,
         Black_Knight_Moves = Knight_Moves - White_Knight_Moves,
         Black_Rook_Moves = Rook_Moves - White_Rook_Moves,
         Black_King_Moves = King_Moves - White_King_Moves,
         Black_Queen_Moves = Queen_Moves - White_Queen_Moves,
         My_Bishop_Moves = if_else(Played_White == 1, White_Bishop_Moves, Black_Bishop_Moves),
         My_Knight_Moves = if_else(Played_White == 1, White_Knight_Moves, Black_Knight_Moves),
         My_Rook_Moves = if_else(Played_White == 1, White_Rook_Moves, Black_Rook_Moves),
         My_King_Moves = if_else(Played_White == 1, White_King_Moves, Black_King_Moves),
         My_Queen_Moves = if_else(Played_White == 1, White_Queen_Moves, Black_Queen_Moves),
         Opponent_Bishop_Moves = if_else(Played_White == 0, White_Bishop_Moves, Black_Bishop_Moves),
         Opponent_Knight_Moves = if_else(Played_White == 0, White_Knight_Moves, Black_Knight_Moves),
         Opponent_Rook_Moves = if_else(Played_White == 0, White_Rook_Moves, Black_Rook_Moves),
         Opponent_King_Moves = if_else(Played_White == 0, White_King_Moves, Black_King_Moves),
         Opponent_Queen_Moves = if_else(Played_White == 0, White_Queen_Moves, Black_Queen_Moves))

## Viewing cleaned and new columns

games %>%
  select(Num_Moves, My_Bishop_Moves, My_Knight_Moves, My_Rook_Moves, My_Queen_Moves, My_King_Moves)

```

#### Creating columns to show number of checks

```{r}
games = games %>%
  mutate(Black_Checks = str_count(Moves, '\\+ [0-9]'),
         Total_Checks = str_count(Moves, '\\+'),
         White_Checks = Total_Checks - Black_Checks,
         My_Checks = if_else(Played_White == 1, White_Checks, Black_Checks),
         Opponent_Checks = if_else(Played_White == 0, White_Checks, Black_Checks))
```

#### Creating columns to show king and queenside castling

```{r}
games = games %>%
  mutate(Black_Kingside_Castled = as.numeric(str_detect(Moves, ' O-O [0-9]')),
         Black_Queenside_Castled = as.numeric(str_detect(Moves, ' O-O-O [0-9]')),
         White_Kingside_Castled = as.numeric(str_detect(Moves, '\\. O-O ')),
         White_Queenside_Castled = as.numeric(str_detect(Moves, '\\. O-O-O ')),
         My_Kingside_Castled = if_else(Played_White == 1, White_Kingside_Castled, Black_Kingside_Castled),
         My_Queenside_Castled = if_else(Played_White == 1, White_Queenside_Castled, Black_Queenside_Castled),
         Opponent_Kingside_Castled = if_else(Played_White == 0, White_Kingside_Castled, Black_Kingside_Castled),
         Opponent_Queenside_Castled = if_else(Played_White == 0, White_Queenside_Castled, Black_Queenside_Castled),
         Kingside_Castled_vs_Queenside_Castled = if_else(My_Kingside_Castled + Opponent_Queenside_Castled == 2, 1, 0),
         Queenside_Castled_vs_Kingside_Castled = if_else(My_Queenside_Castled + Opponent_Kingside_Castled == 2, 1, 0),
         Opposite_Castles = if_else(Kingside_Castled_vs_Queenside_Castled + Queenside_Castled_vs_Kingside_Castled > 0, 1, 0))
```

#### Removing games with low number of moves

```{r}

summary(games$Num_Moves)

hist(x = games$Num_Moves,
     breaks = 20,
     freq = TRUE,
     col = "lightblue",
     xlab = "Number of Moves",
     ylab = "Games",
     main = "Frequency of Games by Number of Moves",
     ylim = c(0,400))

games = games %>%
  filter(Num_Moves > 5)

```

#### Creating columns for rating differences

```{r}
## Creating columns for my and my opponent's ratings and rating differences

games = games %>%
  mutate(My_Rating = if_else(White == "username", White_Rating, Black_Rating),
         Oponnent_Rating = if_else(White == "username", Black_Rating, White_Rating),
         Rating_Diff_vs_Opponent = My_Rating - Oponnent_Rating)

# Creating categories for rating differences

summary(games$Rating_Diff_vs_Opponent)

games = games %>%
  mutate(Rating_Diff_vs_Opponent_Cat = cut(Rating_Diff_vs_Opponent,
                                           breaks = c(-1000, -100, -75, -50, -25, 0, 25, 50, 75, 100, 1000),
                                           labels = c("-100 or Less", "-99 to -75", "-74 to -50", "-49 to -25", "-25 to 0", "1 to 25", "26 to 50", "51 to 75", "76 to 100", "101 or Higher")))
```

#### Creating variables to show result of last game

```{r}
## Sort games from earliest to latest using ID column from earlier

games = games %>%
  arrange(ID)

## Adding columns for previous result and rating changes

games = games %>%
  mutate(My_Rating_Change = if_else(Played_White == 1, White_Rating_Diff, Black_Rating_Diff))

games = games %>%
  mutate(Previous_Result = lag(games$My_Result),
         Previous_Rating_Change = lag(games$My_Rating_Change))

## Viewing new columns

games %>%
  select(ID, My_Result, My_Rating_Change, Previous_Result, Previous_Rating_Change)
```

#### Examining the finalized data frame

```{r}
skim(games)
```
## Data Exploration & Insights

#### On which side of the board do I play better and is there a significant difference in win percentage?

```{r}
## Performance as white

games %>%
  filter(White_or_Black == "White") %>%
  group_by(My_Result_Cat) %>%
  summarize(Games = n()) %>%
  mutate(Percentage = Games/sum(Games)) %>%
  arrange(desc(Percentage))

## Performance as black

games %>%
  filter(White_or_Black == "Black") %>%
  group_by(My_Result_Cat) %>%
  summarize(Games = n()) %>%
  mutate(Percentage = Games/sum(Games)) %>%
  arrange(desc(Percentage))

## Significant difference between wins as black or white?

games.as.white = games %>%
  filter(White_or_Black == "White") %>%
  nrow()

wins.as.white = sum(games$I_Win_As_White)

games.as.black = games %>%
  filter(White_or_Black == "Black") %>%
  nrow()

wins.as.black = sum(games$I_Win_As_Black)

prop.test(x = c(wins.as.white, wins.as.black), n = c(games.as.white, games.as.black))

```
#### Which openings may be under or over-utilized in my repertoire as white?

```{r}
### What are the openings I play most as white and my win percentage in each?

library(ggplot2)
  
## Top 15 openings played as white

games %>%
  filter(Played_White == 1) %>%
  group_by(Opening_Type) %>%
  summarize(Games = n()) %>%
  top_n(n = 15, wt = Games) %>%
  ggplot(mapping = aes(x = reorder(Opening_Type, -Games),
                       y = Games,
                       fill = Opening_Type,
                       label = Games)) +
  geom_col() +
  geom_text(nudge_y = 15, size = 3, fontface = "bold") +
  labs(title = 'Top 15 Openings Played as White',
       x = 'Openings')+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        plot.title = element_text(hjust = 0.5),
        panel.background = element_blank(),
        legend.position = "none")

## Win percentage vs. total games for top 15 openings played as white

library(scales)
library(ggrepel)

games %>%
  filter(Played_White == 1) %>%
  group_by(Opening_Type) %>%
  summarize(Games = n(),
            Wins = sum(I_Win),
            Win.Perc = round((Wins/Games),2)) %>%
  top_n(n = 15, wt = Games) %>%
  ggplot(mapping = aes(x = Games,
                       y = percent(x = Win.Perc, accuracy = 1L),
                       col = Opening_Type,
                       label = Opening_Type)) +
  geom_text_repel(size = 3.25, fontface = "bold") +
  geom_point(col = "white") +
  labs(title = 'Win Percentage in Top 15 Openings Played as White',
       x = 'Games',
       y = 'Win Percentage')+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        plot.title = element_text(hjust = 0.5),
        panel.background = element_blank(),
        legend.position = "none")
```
#### Which openings may be under or over-utilized in my repertoire as black?

```{r}
## Top 15 Openings Played as Black

games %>%
  filter(Played_Black == 1) %>%
  group_by(Opening_Type) %>%
  summarize(Games = n()) %>%
  top_n(n = 15, wt = Games) %>%
  ggplot(mapping = aes(x = reorder(Opening_Type, -Games),
                       y = Games,
                       fill = Opening_Type,
                       label = Games)) +
  geom_col() +
  geom_text(nudge_y = 15, size = 3, fontface = "bold") +
  labs(title = 'Top 15 Openings Played as Black',
       x = 'Openings')+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        plot.title = element_text(hjust = 0.5),
        panel.background = element_blank(),
        legend.position = "none")

## Win Percentage vs. Total Games for Top 15 Openings Played as Black

library(scales)
library(ggrepel)

games %>%
  filter(Played_Black == 1) %>%
  group_by(Opening_Type) %>%
  summarize(Games = n(),
            Wins = sum(I_Win),
            Win.Perc = round((Wins/Games),2)) %>%
  top_n(n = 15, wt = Games) %>%
  ggplot(mapping = aes(x = Games,
                       y = percent(x = Win.Perc, accuracy = 1L),
                       col = Opening_Type,
                       label = Opening_Type)) +
  geom_text_repel(size = 3.25, fontface = "bold") +
  geom_point(col = "white") +
  labs(title = 'Win Percentage in Top 15 Openings Played as Black',
       x = 'Games',
       y = 'Win Percentage')+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        plot.title = element_text(hjust = 0.5),
        panel.background = element_blank(),
        legend.position = "none")
```
#### During which days of the week is my play most effective?

```{r}

# Create data frame to show total number of games by day

day.games = games %>%
  group_by(Day) %>%
  summarize(Total.Games = n())

# Create column to chart using inner join to show win, draw, loss percentage

games %>%
  group_by(Day, My_Result_Cat) %>%
  summarize(Games = n()) %>%
  inner_join(day.games, by = c("Day" = "Day")) %>%
  mutate(Perc = Games/Total.Games) %>%
  ggplot(mapping = aes(x = Day,
                       y = Games,
                       fill = My_Result_Cat,
                       label = percent(x = Perc, accuracy = 1L))) +
  geom_col(position = "stack") +
  geom_text(size = 2.75, position = position_stack(0.5), fontface = "bold") +
  labs(title = 'Results by Day of Week',
       x = 'Day of Week')+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        plot.title = element_text(hjust = 0.5),
        panel.background = element_blank())+
  guides(fill=guide_legend(title='My Result'))+
  scale_fill_manual(values = c("grey", "pink", "cornflowerblue"))
```

#### How much more effective is my play during the weekend and is there a significant difference in win percentage vs. weekdays?

```{r}
# Creating two new variables for weekday and weekend

games$Weekday = as.numeric(str_detect(games$Day, 'Monday|Tuesday|Wednesday|Thursday|Friday'))
games$Weekend = as.numeric(str_detect(games$Day, 'Saturday|Sunday'))

# Results during weekdays

games %>%
  filter(Weekday == 1) %>%
  group_by(My_Result_Cat) %>%
  summarize(Total = n(),
            Percentage = round(Total/sum(games$Weekday),2))

# Results during weekends

games %>%
  filter(Weekend == 1) %>%
  group_by(My_Result_Cat) %>%
  summarize(Total = n(),
            Percentage = round(Total/sum(games$Weekend),2))

# Is there a significant difference in win percentage on weekdays vs. weekends?

prop.test(x = c(824, 482), n = c(sum(games$Weekday), sum(games$Weekend)))

```
#### How does the distribution of piece moves differ between wins and losses?

```{r}

# Create data frame with "longer" data showing piece moves in preparation for visualization

piece.moves = games %>%
  select(My_Result_Cat, My_Knight_Moves, My_Bishop_Moves, My_Rook_Moves, My_Queen_Moves, My_King_Moves) %>%
  filter(My_Result_Cat != "Draw") %>%
  pivot_longer(cols = c("My_Knight_Moves", "My_Bishop_Moves", "My_Rook_Moves", "My_Queen_Moves", "My_King_Moves"), names_to = "Piece_Moves") %>%
  mutate(Result_Piece = paste(My_Result_Cat, Piece_Moves, sep = "_")) %>%
  select(Result_Piece, value) %>%
  rename("Piece_Moves" = "value")

# Replace original piece move names with labelling more appropriate for visualization

piece.moves$Result_Piece = str_replace_all(string = piece.moves$Result_Piece,
                                           c("Loss_My_Knight_Moves" = "Knight Moves in Losses",
                                             "Win_My_Knight_Moves" = "Knight Moves in Wins",
                                             "Loss_My_Bishop_Moves" = "Bishop Moves in Losses",
                                             "Win_My_Bishop_Moves" = "Bishop Moves in Wins",
                                             "Loss_My_Rook_Moves" = "Rook Moves in Losses",
                                             "Win_My_Rook_Moves" = "Rook Moves in Wins",
                                             "Loss_My_Queen_Moves" = "Queen Moves in Losses",
                                             "Win_My_Queen_Moves" = "Queen Moves in Wins",
                                             "Loss_My_King_Moves" = "King Moves in Losses",
                                             "Win_My_King_Moves" = "King Moves in Wins"))

# Converting the new variable to an ordered factor so that the visualization will display a specific order

piece.moves$Result_Piece = factor(piece.moves$Result_Piece,
                                  ordered = TRUE,
                                  levels = c("Knight Moves in Losses", "Knight Moves in Wins", "Bishop Moves in Losses", "Bishop Moves in Wins", "Rook Moves in Losses", "Rook Moves in Wins", "Queen Moves in Losses", "Queen Moves in Wins", "King Moves in Losses", "King Moves in Wins"))

# Creating a multi-box plot visualization of piece moves

ggplot(data = piece.moves,
       mapping = aes(x = Result_Piece, y = Piece_Moves, fill = Result_Piece))+ 
  geom_boxplot(outlier.shape = NA) +
  stat_summary(fun = "mean", geom = "point", size = 1, color = "blue") +
  ylim(c(0, 15)) +
  labs(x = "Pieces & Results",
       y = "Piece Moves",
       title = "Distribution of Piece Moves for Wins vs. Losses") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        plot.title = element_text(hjust = 0.5),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"),
        legend.position = "none")+
  scale_fill_manual(values = c("lightblue2", "lightblue2","lightgreen", "lightgreen","pink", "pink","khaki", "khaki","grey", "grey"))

# Is the difference between pieces moves in wins vs. losses statistically significant?

# First, I create a data frame with just piece moves and results, filtered to only wins and losses so that the variable only has two levels.

pieces.results = games %>%
  select(My_Result_Cat, My_Bishop_Moves, My_Knight_Moves, My_Rook_Moves, My_Queen_Moves, My_King_Moves) %>%
  filter(My_Result_Cat != "Draw")

# Employ two-sample t-tests

knight = t.test(My_Bishop_Moves~My_Result_Cat, data = pieces.results)
bishop = t.test(My_Knight_Moves~My_Result_Cat, data = pieces.results)
rook = t.test(My_Rook_Moves~My_Result_Cat, data = pieces.results)
queen = t.test(My_Queen_Moves~My_Result_Cat, data = pieces.results)
king = t.test(My_King_Moves~My_Result_Cat, data = pieces.results)

# Assign values for defining means and differences between means

knight.losses = piece.moves %>% filter(Result_Piece == "Knight Moves in Losses")
mean.knight.losses = round(mean(knight.losses$Piece_Moves),2)
knight.wins = piece.moves %>% filter(Result_Piece == "Knight Moves in Wins")
mean.knight.wins = round(mean(knight.wins$Piece_Moves), 2)
knight.wins.vs.losses = round((mean.knight.wins - mean.knight.losses),2)

bishop.losses = piece.moves %>% filter(Result_Piece == "Bishop Moves in Losses")
mean.bishop.losses = round(mean(bishop.losses$Piece_Moves),2)
bishop.wins = piece.moves %>% filter(Result_Piece == "Bishop Moves in Wins")
mean.bishop.wins = round(mean(bishop.wins$Piece_Moves),2)
bishop.wins.vs.losses = round((mean.bishop.wins - mean.bishop.losses),2)

rook.losses = piece.moves %>% filter(Result_Piece == "Rook Moves in Losses")
mean.rook.losses = round(mean(rook.losses$Piece_Moves), 2)
rook.wins = piece.moves %>% filter(Result_Piece == "Rook Moves in Wins")
mean.rook.wins = round(mean(rook.wins$Piece_Moves), 2)
rook.wins.vs.losses = round((mean.rook.wins - mean.rook.losses),2)

queen.losses = piece.moves %>% filter(Result_Piece == "Queen Moves in Losses")
mean.queen.losses = round(mean(queen.losses$Piece_Moves), 2)
queen.wins = piece.moves %>% filter(Result_Piece == "Queen Moves in Wins")
mean.queen.wins = round(mean(queen.wins$Piece_Moves), 2)
queen.wins.vs.losses = round((mean.queen.wins - mean.queen.losses), 2)

king.losses = piece.moves %>% filter(Result_Piece == "King Moves in Losses")
mean.king.losses = round(mean(king.losses$Piece_Moves), 2)
king.wins = piece.moves %>% filter(Result_Piece == "King Moves in Wins")
mean.king.wins = round(mean(king.wins$Piece_Moves), 2)
king.wins.vs.losses = round((mean.king.wins - mean.king.losses),2)

# Create data frame to identify which are statistically signficant

Pieces = c("Knight", "Bishop", "Rook", "Queen", "King")
Avg_Moves_in_Wins = c(mean.knight.wins, mean.bishop.wins, mean.rook.wins, mean.queen.wins, mean.king.wins)
Avg_Moves_in_Losses = c(mean.knight.losses, mean.bishop.losses, mean.rook.losses, mean.queen.losses, mean.king.losses)
Difference = Avg_Moves_in_Wins - Avg_Moves_in_Losses
P_Values = c(round(knight$p.value,3), round(bishop$p.value, 3), round(rook$p.value, 3), round(queen$p.value, 3), round(king$p.value, 3))

Pieces_P_Values = data.frame(Pieces, Avg_Moves_in_Wins, Avg_Moves_in_Losses, Difference, P_Values) %>%
  mutate(Significant = P_Values<=0.05); Pieces_P_Values
```

#### Does the number of checks by my opponent or myself differ between wins and losses, and if so, is there a significant difference between the two?

```{r}

wins = games %>%
  filter(My_Result_Cat == "Win")

losses = games %>%
  filter(My_Result_Cat == "Loss")

t.test(wins$My_Checks, losses$My_Checks)
t.test(wins$Opponent_Checks, losses$Opponent_Checks)

```

#### How common are repeated results, and are winning streaks or losing streaks more common?

```{r}
# Percentage of games with repeated results, repeated wins and repeated losses

games %>%
  select(ID, My_Result, Previous_Result) %>%
  filter(!is.na(Previous_Result)) %>%
  mutate(Repeated_Result = if_else(My_Result == Previous_Result, 1, 0),
         Repeated_Win = if_else(My_Result + Previous_Result == 2, 1, 0),
         Repeated_Loss = if_else(My_Result + Previous_Result == 0, 1, 0)) %>%
  summarize(Repeated_Results_All_Games = round((sum(Repeated_Result)/n()),2),
            Repeated_Wins_All_Games = round(sum(Repeated_Win)/n(),2),
            Repeated_Losses_All_Games = round(sum(Repeated_Loss)/n(),2))

# Percentage of wins which were repeated wins and percentage of losses which were repeat losses

games %>%
  select(ID, My_Result, Previous_Result) %>%
  filter(!is.na(Previous_Result)) %>%
  mutate(Repeated_Result = if_else(My_Result == Previous_Result, 1, 0),
         Repeated_Win = if_else(My_Result + Previous_Result == 2, 1, 0),
         Repeated_Loss = if_else(My_Result + Previous_Result == 0, 1, 0)) %>%
  summarize(Repeated_Wins_Winning_Games = round(sum(Repeated_Win)/sum(games$I_Win),2),
            Repeated_Losses_Losing_Games = round(sum(Repeated_Loss)/sum(games$I_Lose),2))

# Correlation testing

library(rstatix)

cor_test(data = games,
         vars = 'My_Result',
         vars2 = c('Previous_Result', 'Previous_Rating_Change'),
         use = "complete.obs")

# Statistical significance testing

games %>%
  select(ID, My_Result, Previous_Result) %>%
  filter(!is.na(Previous_Result)) %>%
  mutate(Repeated_Win = if_else(My_Result + Previous_Result == 2, 1, 0),
         Repeated_Loss = if_else(My_Result + Previous_Result == 0, 1, 0)) %>%
  summarize(Repeated_Wins_Total = sum(Repeated_Win),
            Repeated_Losses_Total = sum(Repeated_Loss))

prop.test(x = c(663, 564), n = c(sum(games$I_Win), sum(games$I_Lose)))
```

#### Rating fluctuations

```{r}
blitz.games = games %>%
  filter(str_detect(Game_Type, 'Blitz'),
         Month_Year != 'Dec 2022')

boxplot(My_Rating ~ Month_Year, data = blitz.games,
        xlab = "Month & Year",
        ylab = "Rating",
        main = "Boxplots of Blitz Ratings by Month",
        col = "lightblue",
        outline = FALSE)

summary(blitz.games$My_Rating)

### Seeing months where I had the biggest differences in ratings

blitz.games %>%
  group_by(Month_Year) %>%
  summarize(Rating_Range = max(My_Rating) - min(My_Rating)) %>%
  ggplot(mapping = aes(x = Month_Year,
                     y = Rating_Range,
                     label = Rating_Range)) +
  geom_col(fill = "cornflowerblue") +
  ylim(0, 150) +
  geom_text(nudge_y = -10, colour = "white", size = 3) +
  labs(title = 'Rating Range by Month',
       x = 'Month & Year',
       y = 'Rating Range')+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        plot.title = element_text(hjust = 0.5),
        panel.background = element_blank(),
        legend.position = "none")
  
```
## Predictive Modeling & Inferences

#### Feature Selection: Examining bivariate correlations to identify sources of collinearity

```{r}

## Creating data frame with features of interest to examine correlations

games.1 = games %>%
  select(Hourly_Blitz_Arena, Hourly_SuperBlitz_Arena, Rated_Blitz, Rated_Bullet, Rated_Rapid, Rated_Bullet,
         Weekend, Weekday,
         Played_White, Played_Black,
         Num_Moves,
         My_Bishop_Moves, My_Knight_Moves, My_Rook_Moves, My_Queen_Moves, My_King_Moves,
         Opponent_Bishop_Moves, Opponent_Knight_Moves, Opponent_Rook_Moves, Opponent_Queen_Moves, Opponent_King_Moves,
         My_Checks, Opponent_Checks,
         My_Rating, Oponnent_Rating,
         Previous_Result, Previous_Rating_Change,
         I_Win)

## Splitting data into train and test

library(caTools)
set.seed(100)
split = sample.split(games.1$I_Win, SplitRatio = 0.7)
train = games.1[split,]
test = games.1[!split,]

str(train)

## Examining bivariate correlations

library(rstatix)

cor = cor_test(data = train, use = 'complete.obs') %>%
  data.frame() %>%
  filter(var1 != var2)

## Examining correlations with winning binary indicator

cor %>%
  filter(var1 == 'I_Win') %>%
  mutate(significant = p <= 0.05) %>%
  select(var1, var2, cor, significant) %>%
  filter(significant == 'TRUE') %>%
  arrange(desc(cor)) %>%
  View()

## Filtering to strong correlation (higher than 0.7 or or lower than -0.7)

cor %>%
  filter(abs(cor) >= 0.7,
         !duplicated(statistic)) %>%
  arrange(desc(cor))

## Examining correlations with binary indicator for winning the games

cor_test(data = train,
         vars = 'I_Win',
         vars2 = c('Previous_Result', 'Previous_Rating_Change', 'Num_Moves',
                   'Opponent_King_Moves', 'My_King_Moves', 'My_Checks',
                   'Rated_Bullet', 'My_Rating', 'Weekend',
                   'Weekday', 'Played_White', 'Played_Black'),
         use = 'complete.obs') %>%
  arrange(desc(cor)) %>%
  mutate(significant = p <= 0.05) %>%
  select(var1, var2, cor, significant) %>%
  View()

## Removing redundant features and those which have smaller correlation with I_Win

train = train %>%
  select(-Previous_Rating_Change, -Num_Moves, -Opponent_King_Moves, -Rated_Bullet, -Weekend, -Played_White) %>%
  filter(!is.na(Previous_Result))
```

#### Logistic regression model

```{r}

## Training model

glm1 = glm(I_Win ~ ., data = train, family = "binomial"(link = "logit"))

## Creating predictions for train data

train.pred = predict(glm1, data = train, type = 'response')

## Comparing actual results from train data vs. predicted results

actual.vs.pred = data.frame(Wins = train$I_Win, Probability = train.pred, Predictions = as.numeric(train.pred >= 0.5)) %>%
  mutate(Correct = as.numeric(Wins == Predictions))

actual.vs.pred[1:10,]

## Calculating accuracy

sum(actual.vs.pred$Correct)/nrow(actual.vs.pred) # 66.3% accuracy

## Creating predictions for test data

test.pred = predict(glm1, newdata = test, type = 'response')

## Comparing actual results from test data vs. predicted results

actual.vs.pred = data.frame(Wins = test$I_Win, Probability = test.pred, Predictions = as.numeric(test.pred > 0.5)) %>%
  mutate(Correct = as.numeric(Wins == Predictions))

actual.vs.pred[1:10,]

## Calculating accuracy

sum(actual.vs.pred$Correct)/nrow(actual.vs.pred) # 64.4% accuracy

## Inspecting model

summary(glm1)

## Odds ratio

format((exp(glm1$coefficients)-1)*100, scientific = FALSE)

```

#### Decision tree model

```{r}
## Loading libraries

library(rpart); library(rpart.plot)

## Creating model based on train data

tree1 = rpart(I_Win ~ ., data = train, method = "class")

## Prediction accuracy on train data

pred1.prob = predict(tree1, data = train, type = "prob")
pred1.class = predict(tree1, data = train, type = "class")
train.results = train$I_Win

data.frame(train.results, pred1.prob, pred1.class) %>%
  mutate(Correct_Prediction = if_else(train.results == pred1.class, 1, 0)) %>%
  summarize(sum(Correct_Prediction)/n()) #65%

## Prediction accuracy on test data

pred1.prob = predict(tree1, newdata = test, type = "prob")
pred1.class = predict(tree1, newdata = test, type = "class")
test.results = test$I_Win

data.frame(test.results, pred1.prob, pred1.class) %>%
  mutate(Correct_Prediction = if_else(test.results == pred1.class, 1, 0)) %>%
  summarize(sum(Correct_Prediction)/n()) #63%

## Inspecting model and variable importance

summary(tree1)

rpart.plot(tree1)

tree1$variable.importance

```

#### Random forest model

```{r}

## Loading library and setting seed

library(randomForest)
set.seed(100)

## Creating model based on train data

forest1 = randomForest(as.factor(I_Win)~., data = train, ntree = 1000)

## Predictions using train data

pred_forest_prob = predict(forest1, type = "prob")
pred_forest_class = predict(forest1, type = "class")

data.frame(train$I_Win, pred_forest_prob, pred_forest_class) %>%
  mutate(correct_prediction = if_else(train.I_Win == pred_forest_class, 1, 0)) %>%
  summarize(sum(correct_prediction)/n()) #64%

## Predictions using test data

pred_forest_prob = predict(forest1, newdata = test, type = "prob")
pred_forest_class = predict(forest1, newdata = test, type = "class")

data.frame(test$I_Win, pred_forest_prob, pred_forest_class) %>%
  mutate(correct_prediction = if_else(test.I_Win == pred_forest_class, 1, 0)) %>%
  summarize(sum(correct_prediction)/n()) #62%

## Inspecting variable importance

forest1$importance

varImpPlot(forest1)
```

```


